Syntax and Semantics of Abstract Binding Trees

### 2015-12-02 paper.tex
Rework the treatment of operators
@freebroccolo What do you think of this?

\newcommand\FLift[2]{
  \Operators\langle#1%
  \ifthenelse{\equal{#2}{}}{}{, #2}\rangle%
}

@@

\newcommand\Sheaves[1]{\mathbf{Sh}(#1)}

@@

Intuitively, when $\Upsilon$ supports $\Member{m}{X(\Upsilon')}$, we would
expect that we can work backward to a unique $\Of{m'}{\Upsilon}$ such that
$\IsEq{m'\varrho}{m}$; whilst this is not in general the case for presheaves
$X$, it is precisely the sheaf condition for covariant presheafs on $\SCtx$
under the atomic topology \cite[p.~126]{mac-lane-moerdijk:1992}. The sheaf condition, then, ensures that the notion
of support is well-behaved.

@@

\begin{remark}
  The presentation we have given above is related both to the Schanuel topos
  and, equivalently, Pitts' category of nominal sets \cite{fiore-staton:2006}; we differ only in
  discussing covariant sheaves on the category of symbol contexts $\SCtx$, as
  opposed to covariant sheaves on finite sets of symbols $\SymSets$ (i.e.\
  unisorted contexts).
\end{remark}

@@

Fix a sheaf of operators
$\Of{\Operators}{\Sheaves{\OpCat{\left(\SCtx\right)}\times\Discrete{\Arities}}}$
such that the arrows in $\SCtx$ lift to renamings of operators' parameters.
Together, $\Sorts$ and $\Operators$ form a \emph{signature}
$\Define{\Sigma}{\Pair{\Sorts}{\Operators}}$.

By the Grothendieck construction
$\Iso{\Grothendieck{}{\Operators}}{\Comma{\mathbb{1}}{\Operators}}$ we have a
category of elements of $\Operators$ where objects are triples
$\Pair{\Pair{\Upsilon}{a}}{\vartheta}$, interpreted as
$\Member{\vartheta}{\Operators\Pair{\Upsilon}{a}}$, and morphisms are paired
diagrams:
\[
  \begin{tikzcd}[cramped, ampersand replacement = \&]
    \IMode{\Pair{\Upsilon}{a}}
      \arrow[r, swap, "\IMode{\Pair{\rho}{f}}"]
\&
    \IMode{\Pair{\Upsilon'}{a'}}
  \end{tikzcd}
\qquad
  \begin{tikzcd}[cramped, sep = small, ampersand replacement = \&]
    \null
\&
    \IMode{\mathbb{1}}
      \arrow[dl, swap, "\IMode{\vartheta}"]
      \arrow[dr, "\IMode{\vartheta'}"]
\&
    \null
\\
    \IMode{\Operators\Pair{\Upsilon}{a}}
      \arrow[rr, swap, "\IMode{\Operators\Pair{\rho}{f}}"]
\&
    \null
\&
    \IMode{\Operators\Pair{\Upsilon'}{a'}}
  \end{tikzcd}
\]

We will write $\IsOperator{\Upsilon}{\vartheta}{a}$ in case
$\Member{\vartheta}{\Operators\Pair{\Upsilon}{a}}$ and note that this judgment
enjoys the structural properties of weakening and exchange via functoriality of
$\Operators$. Operators $\vartheta_{0}, \ldots, \vartheta_{n}$ are defined by
specifying the fibers of
$\Of{\pi_{\Operators}}{\Grothendieck{}{\Operators}\ra\SCtx\times\Discrete{\Arities}}$
in which they reside, such that the glueing condition is satisfied:
\begin{alignat*}{3}
  \null
&
  \IMode{{\pi_{\Operators}}^{-1}{\Pair{\Upsilon}{a}}}
&&
  \,\mathrel{\ {{:}}}\,
&&
  \IMode{\SCtx\times\Discrete{\Arities}\ra\Sets}
\\
  \null
&
  \IMode{{\pi_{\Operators}}^{-1}{\Pair{\Upsilon}{a}}}
&&
  \,=\,\
&&
    \OMode{
      \MkSet{
        \vartheta\in\Pair{\Upsilon}{a}
        \mid{\pi_{\Operators}}{\Pair{\Pair{\Upsilon}{a}}{\vartheta}} =
        \Pair{\Upsilon}{a}
      }
    }
\end{alignat*}

@@

For instance, consider the $\lambda$-calculus with a single sort, $\SortExp$; we
give its signature $\Sigma_{\lambda}$ by asserting the following about its
operators:
\begin{gather*}
  \IsOperator{\Upsilon}{\mathtt{lam}}{
    \MkArity{
      \MkValence{}{\SortExp}{\SortExp}
    }{\SortExp}
  }
\\
  \IsOperator{\Upsilon}{\mathtt{ap}}{
    \MkArity{
      \MkValence{}{}{\SortExp},
      \MkValence{}{}{\SortExp}
    }{\SortExp}
  }
\end{gather*}
These rules correspond to objects of $\Grothendieck{}{{\Operators}_{\lambda}}$
and we collect their fibers into sets of operators:
\begin{alignat*}{5}
  \IMode{{\pi_{{\Operators}_{\lambda}}}^{-1}}
&
  \IMode{
    \Pair{
      \Upsilon
    }{
      \MkArity{
        \MkValence{}{\SortExp}{\SortExp}
      }{\SortExp}
    }
  }
\quad
&&
  =
&&
\quad
  \OMode{\MkSet{\mathtt{lam}}}
\\
  \IMode{{\pi_{{\Operators}_{\lambda}}}^{-1}}
&
  \IMode{
    \Pair{
      \Upsilon
    }{
      \MkArity{
        \MkValence{}{}{\SortExp},
        \MkValence{}{}{\SortExp}
      }{\SortExp}
    }
  }
\quad
&&
  =
&&
\quad
  \OMode{\MkSet{\mathtt{ap}}}
\\
  \IMode{{\pi_{{\Sigma}_{\lambda}}}^{-1}}
&
  \triangleq\IMode{
    \bigcup_{\Pair{\Upsilon}{a}}{
      {\pi_{{\Operators}_{\lambda}}}^{-1}
      \Pair{\Upsilon}{a}
    }
  }
\quad
&&
  =
&&
\quad
  \OMode{
    \MkSet{\mathtt{lam},\ \mathtt{ap}}
  }
\\
\end{alignat*}

So far, we have made no use of symbols and parameters; however, consider the
extension of the calculus with assignables (references):

\begin{gather*}
  \IsOperator{\Upsilon}{\mathtt{decl}}{
    \MkArity{
      \MkValence{}{}{\SortExp},
      \MkValence{\SortExp}{}{\SortExp}
    }{
      \SortExp
    }
  }
\\
  \IsOperator{\Upsilon,u:\SortExp}{\mathtt{get}[u]}{
    \MkArity{}{\SortExp}
  }
\\
  \IsOperator{\Upsilon,u:\SortExp}{\mathtt{set}[u]}{
    \MkArity{
      \MkValence{}{}{\SortExp}
    }{
      \SortExp
    }
  }
\end{gather*}

Declaring a new assignable consists in providing an initial value, and an
expression binding a symbol (which shall represent the assignable in scope).
Note that the functoriality of $\Operators$ guarantees for any renaming
$\Of{\varrho}{\Upsilon\inj\Upsilon'}$ and $\Of{f}{a\ra a'}$, a lifted map
$\Of{\FLift{\varrho}{f}}{\Operators\Pair{\Upsilon}{a}\inj\Operators\Pair{\Upsilon'}{a'}}$
such that $\IsOperator{\Upsilon'}{\FLift{\varrho}{f}(\vartheta)}{a'}$ when
$\IsOperator{\Upsilon}{\vartheta}{a}$. In particular, the renaming
$\Upsilon,u\mapsto\Upsilon,v$ shall take $\mathtt{get}[u]$ to
$\mathtt{get}[v]$. Because the category $\Discrete{\Arities}$ is discrete and
has only identity arrows, through an abuse of notation we will often write
$\vartheta\varrho$ to mean $\FLift{\varrho}{\ArrId{}}(\vartheta)$.

@@

Interpretation of operator applications is the most complicated. Recall that,
unlike in standard treatments of universal algebra, our operators are indexed by
symbol collections; therefore, operators must pass through suitable renamings in
order to be used in the interpretation. Let us begin by constructing for each
operator $\IsOperator{\Upsilon}{\vartheta}{a}$ the morphism
$\SemBrackets{\vartheta}_P$ which shall rename the parameters of the operator
using the environment:
\[
  \begin{tikzcd}[cramped, ampersand replacement = \&]
    |[alias = Src]|
    {\IMode{
      {\SemBrackets{
        \Theta\triangleright\Upsilon\parallel\Gamma
      }}_{P}
    }}
      \arrow[
        , r
        , swap
        , "\IMode{\pi_2}"
      ]
\&
    {\IMode{\SymPsh^{\{\Upsilon\}}}}
      \arrow[
        , r
        , swap
        , "\IMode{\FLift{\pi_{(-)}}{\ArrId{}}}"
      ]
\&[1.5em]
    |[alias = Tgt]|
    {\IMode{\Operators_a}}
      \arrow[
        , from = Src
        , to = Tgt
        , bend left
        , dashed
        , "\OMode{\SemBrackets{\vartheta}_{P}}"
      ]
  \end{tikzcd}
\]

>>>>

@@

\newcommand\Sheaves[1]{\mathbf{Sh}\left(#1\right)}
\newcommand\Presheaves[1]{\mathbf{Psh}\left(#1\right)}
\newcommand\Sheafify[1]{#1^{\#}}
\newcommand\ConstPsh[1]{\Delta[#1]}
\newcommand\ConstSh[1]{\Delta^\#[#1]}
\newcommand\SCtxSite{\SymSets[\Sorts]}

@@

Intuitively, when $\Upsilon$ supports $\Member{m}{X(\Upsilon')}$, we would
expect that we can work backward to a unique $\Of{m'}{\Upsilon}$ such that
$\IsEq{m'\varrho}{m}$; whilst this is not in general the case for presheaves
$X$, it is precisely the sheaf condition for covariant presheaves on $\SCtx$
under the atomic topology \cite[p.~126]{mac-lane-moerdijk:1992}. The sheaf
condition, then, ensures that the notion of support is well-behaved.

@@

Going forward, we will write $\SCtxSite$ for the Grothendieck site
$\Pair{\OpCat{\left(\SCtx\right)}}{A}$, where $A$ is the atomic coverage (sc.\
every non-empty family covers).

\begin{remark}
  The presentation we have given above is related both to the Schanuel topos
  and, equivalently, Pitts' category of nominal sets \cite{fiore-staton:2006}; we differ only in
  discussing sheaves on $\SCtxSite$, as
  opposed to sheaves on $\OpCat{\SymSets}$ (i.e.\ unisorted contexts).
\end{remark}

\subsubsection{Constructions on sheaves}

For any Grothendieck site $\Match{\mathfrak{S}}{\Pair{\mathcal{C}}{J}}$, the
inclusion $\Of{\iota}{\Sheaves{\mathfrak{S}}\ra\Presheaves{\mathcal{C}}}$ of
sheaves into presheaves has a left adjoint
$\IMode{\Sheafify{(-)}}\dashv\IMode{\iota}$, which takes any presheaf
$\Of{X}{\Presheaves{\mathcal{C}}}$ to a unique sheaf
$\Of{\Sheafify{X}}{\Sheaves{\mathfrak{S}}}$, called the \emph{sheafification}
of $X$.

Every category of sheaves on a site gives rise to a topos, which equips us with
a number of standard constructions, including (among other things) the disjoint
union of sheaves $X\oplus Y$, the product of sheaves $X\otimes Y$, and the terminal sheaf $\mathbb{1}$.

Returning to the site $\SCtxSite$, we will write $\mathbf{S}_\tau$ for the sheaf
of symbols of sort $\tau$, a subobject of the Yoneda embedding of the empty
symbol context, $\Yoneda{\cdot}$.

@@

Fix a family of sheaves of operators
$\Member{\Operators_a}{\Sheaves{\SCtxSite}}$, indexed by arities
$\Member{a}{\Arities}$; for each $\Operators_a$, arrows in $\SCtx$ will lift to
renamings in operators' symbolic parameters. Together, $\Sorts$ and $\Operators$ are said to form a \emph{signature}
$\Define{\Sigma}{\Pair{\Sorts}{\Operators}}$.

We will write $\IsOperator{\Upsilon}{\vartheta}{a}$ in case
$\Member{\vartheta}{\Operators_a(\Upsilon)}$; note that this judgment enjoys
the structural properties of weakening and exchange via functoriality of
$\Operators$. An operator signature is defined by specifying, for each arity
$a$, the sheaf $\Operators_a$, whose ``elements'' are the operators of arity
$a$.

@@

For instance, consider a $\lambda$-calculus with a single sort, $\SortExp$; we
give its signature $\Sigma_{\lambda}$ by asserting the following about its
operators:
\begin{gather*}
  \IsOperator{\Upsilon}{\mathtt{lam}}{
    \MkArity{
      \MkValence{}{\SortExp}{\SortExp}
    }{\SortExp}
  }\\
  \IsOperator{\Upsilon}{\mathtt{fix}}{
    \MkArity{
      \MkValence{}{\SortExp}{\SortExp}
    }{\SortExp}
  }\\
  \IsOperator{\Upsilon}{\mathtt{ap}}{
    \MkArity{
      \MkValence{}{}{\SortExp},
      \MkValence{}{}{\SortExp}
    }{\SortExp}
  }
\end{gather*}

These rules correspond to the following definition of $\Operators$:
\begin{align*}
  \ADefine{
    \Operators_{
       \MkArity{
        \MkValence{}{\SortExp}{\SortExp}
      }{\SortExp}
    }
  }{
    \mathbb{1}\oplus\mathbb{1}
  }
  \\
  %
  \ADefine{
    \Operators_{
      \MkArity{
        \MkValence{}{}{\SortExp},
        \MkValence{}{}{\SortExp}
      }{\SortExp}
    }
  }{
    \mathbb{1}
  }
\end{align*}

So far, we have made no use of symbols and parameters; however, consider the
extension of the calculus with assignables (references):
%
\begin{gather*}
  \IsOperator{\Upsilon}{\mathtt{decl}}{
    \MkArity{
      \MkValence{}{}{\SortExp},
      \MkValence{\SortExp}{}{\SortExp}
    }{
      \SortExp
    }
  }
\\
  \IsOperator{\Upsilon,u:\SortExp}{\mathtt{get}[u]}{
    \MkArity{}{\SortExp}
  }
\\
  \IsOperator{\Upsilon,u:\SortExp}{\mathtt{set}[u]}{
    \MkArity{
      \MkValence{}{}{\SortExp}
    }{
      \SortExp
    }
  }
\end{gather*}

These rules correspond to the following family of sheaves:
\begin{align*}
  \ADefine{
    \Operators_{
      \MkArity{
        \MkValence{}{}{\SortExp},
        \MkValence{\SortExp}{}{\SortExp}
      }{
        \SortExp
      }
    }
  }{
    \mathbb{1}
  }\\
  \ADefine{
    \Operators_{
      \MkArity{}{\SortExp}
    }
  }{
    \mathbf{S}_\SortExp
  }
\end{align*}

Declaring a new assignable consists in providing an initial value, and an
expression binding a symbol (which shall represent the assignable in scope).
Note that the functoriality of $\Operators$ guarantees for any renaming
$\Of{\varrho}{\Upsilon\inj\Upsilon'}$, a family of lifted maps
$\Of{\Operators_a(\varrho)}{\Operators_a(\Upsilon)\inj\Operators_a(\Upsilon')}$
natural in $a$, such that
$\IsOperator{\Upsilon'}{\Operators_a(\varrho)(\vartheta)}{a}$ when
$\IsOperator{\Upsilon}{\vartheta}{a}$. In particular, the renaming
$\Upsilon,u\mapsto\Upsilon,v$ shall take $\mathtt{get}[u]$ to
$\mathtt{get}[v]$. Going forward, we will write $\vartheta\varrho$ for
$\Operators_a(\varrho)$.

@@

Interpretation of operator applications is the most complicated. Recall that,
unlike in standard treatments of universal algebra, our operators are indexed by
symbol collections; therefore, operators must pass through suitable renamings in
order to be used in the interpretation. Let us begin by constructing for each
operator $\IsOperator{\Upsilon}{\vartheta}{a}$ the morphism
$\SemBrackets{\vartheta}_P$ which shall rename the parameters of the operator
using the environment:
\[
  \begin{tikzcd}[cramped, ampersand replacement = \&, sep = huge]
    |[alias = Src]|
    {\IMode{
      {\SemBrackets{
        \Theta\triangleright\Upsilon\parallel\Gamma
      }}_{P}
    }}
      \arrow[
        , r
        , swap
        , "\IMode{\pi_2}"
      ]
\&
    {\IMode{\SymPsh^{\{\Upsilon\}}}}
      \arrow[
        , r
        , swap
        , "\IMode{\Operators_a(\pi_{(-)})(\vartheta)}"
      ]
\&[1.5em]
    |[alias = Tgt]|
    {\IMode{\Operators_a}}
      \arrow[
        , from = Src
        , to = Tgt
        , bend left
        , dashed
        , "\OMode{\SemBrackets{\vartheta}_{P}}"
      ]
  \end{tikzcd}
\]
